On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-28 paper/results2d.tex
MKL version 3-ing

Eq. \ref{inviscid} is a cubic equation in $S$. The 
Routh-Hurwitz criteria imply that stability is ensured if 
\begin{align}\label{stable_condition}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. 
%gamma/theta condition is related to some kind of thermal instability
%oscilitory convection?? 
%second condition concerning Q is associated with self-g 
A third criterion, $2(2-q)\gamma Q^2>1$, is formally required, but this
is implied by Eq. \ref{stable_condition}. 
Notice these conditions do not actually depend on the cooling time.

%As perturbations are cooled to lower temperatures
%(decreasing $\theta$), the minimum $Q$ required for stability
%increases.

If we define $Q_\mathrm{eff}=\sqrt{\theta}Q$, the second condition in
Eq. \ref{stable_condition} is $Q_\mathrm{eff}>1$ (for
the Keplerian case). In other words, if one  
insists on applying a Toomre condition to a cooling disk, then the 
Toomre parameter should be calculated with the floor
temperature. %rather than the actual temperature 
At fixed $Q$ this stability condition will fail for sufficiently small 
$\theta$, i.e. as perturbations are allowed to cool to lower
temperatures.

@@

Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p|K| =
0$ and $S = S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
\begin{align}\label{kstar}
%  |K_*| = \frac{1}{\gamma \qtwo} \left(1 + \frac{1}{\beta S_*}\right). 
  |K_*| = \frac{1+\beta S_*}{Q\left(\theta + \gamma \beta S_*\right)}
\end{align}
%Eq. \ref{kstar} describes how the most unstable  mode is affected by
%cooling ($\beta$) and external heating ($\theta$). 
Inserting this into Eq. \ref{inviscid}, we find the maximum growth
rate satisfies
\begin{align}\label{inviscid_max}
%  S_*^3 + \left[2(2-q)  - \frac{1}{\gamma \qtwo^2}\right]S_* - \frac{1}{\gamma
%    \qtwo^2 \beta} = 0.
  S_*^2 = \frac{1+\beta S_*}{Q^2\left(\theta + \beta\gamma S_*\right)}
  - 2(2-q).
\end{align}
Eq. \ref{kstar}---\ref{inviscid_max} imply $\p_\beta|K_*|,\,\p_\beta
S_*<0 $ for $\gamma>\theta$. As cooling becomes more rapid the maximum 
growth rate increases, and the most unstable wavelength decreases. 
%show dK*/dbeta < 0 
%by diff K* wrt beta and diff S* wrt beta from the equations
%and elim dS*/d beta

If $|S_*|\ll1$, Eq. \ref{inviscid_max} gives the simple solution 
\begin{align}
 \beta S_*\simeq \frac{1-2(2-q)Q^2\theta}{2(2-q)Q^2\gamma-1},
\end{align}
implying that the system can still be formally unstable for arbitarily 
long cooling times. {\bf optional equation?}

@@

\subsubsection{$\theta = 0$}\label{theta0}
For beta cooling with $\theta=0$, the disk is unconditionally unstable
for finite $Q$. {\bf KMK some caveating about scales}  The $\beta\to0$ limit corresponds to a 
pressureless disk (not merely isothermal).

%for definiteness 
Let us consider a disk with 
\begin{align}\label{qcrit_def}
  \qtwo = \frac{1}{\sqrt{2\gamma(2-q)}}\equiv Q_{\mathrm{crit}},
\end{align} 
{\bf KMK $\gamma = \theta$ from eq. 47? I'm missing something}
so that it is marginally stable in the absence of cooling.  
How does finite cooling destabilize the disk?  
Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} and setting
$\theta=0$, we find 
\begin{align}\label{sstar}
  S_*^3 = \frac{1}{\gamma Q_\mathrm{crit}^2 \beta}. 
\end{align}
The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
increases with decreasing $\beta$. Instability formally occurs for
arbitrarily long but finite cooling times. 
%Instability does not require the
%cooling time to be below some critical value.

Nevertheless, we can define a characteristic cooling
time $\beta_*$ as that which removes pressure support against
self-gravity over the natural lengthscale in the problem, the
scale-height $H$. We thus set $|K_*|=1$ and find, for the Keplerian
disk,

\begin{align}\label{betastar}
  \beta_* = \frac{1}{\left(\sqrt{\gamma} - 1\right)^{3/2}}. 
\end{align}
This equation reproduces remarkably similar values 
of the cooling times below which 
%dynamical disk 
%fragmentation occurs in numerical simulations
numerical simulations show dynamical disk fragmentation 
\citep{gammie01,rice05,rice11}. These simulations employ the same beta  
cooling prescription with $\theta=0$, and determine the fragmentation
boundary, $\beta_c$, as a function of the adiabatic index $\gamma$.  
Table \ref{bstar_compare} shows rough agreement between $\beta_*$ and 
$\beta_c$.

@@

\subsubsection{$\theta = 1$}\label{theta1}
Standard beta cooling with $\theta=1$ corresponds to 
`thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case 
$\beta\to 0$ corresponds to an isothermal disk and $\beta
\to \infty$ corresponds to an adiabatic disk. (These two limits are
equivalent for $\gamma=1$.) The most unstable wavenumber and growth
rates vary weakly with $\beta$.   {\bf KMK: summarize boring result? not sure 
what this simplifies to in comparison to other limits?}.

>>>>

Eq. \ref{inviscid} is a cubic equation in $S$. The 
Routh-Hurwitz criteria imply that stability is ensured if 
\begin{align}\label{stable_condition}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. 
%gamma/theta condition is related to some kind of thermal instability
%oscilitory convection?? 
%second condition concerning Q is associated with self-g 
A third criterion, $2(2-q)\gamma Q^2>1$, is formally required, but this
is implied by Eq. \ref{stable_condition}. 
Notice these conditions do not actually depend on the cooling time. 
At fixed $Q$, the second stability condition eventually fails for
decreasing $\theta$, i.e. if perturbations are allowed
to cool to sufficiently low temperatures.

@@

Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p|K| =
0$ and $S = S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
\begin{align}\label{kstar}
%  |K_*| = \frac{1}{\gamma \qtwo} \left(1 + \frac{1}{\beta S_*}\right). 
  |K_*| = \frac{1+\beta S_*}{Q\left(\theta + \gamma \beta S_*\right)}
\end{align}
%Eq. \ref{kstar} describes how the most unstable  mode is affected by
%cooling ($\beta$) and external heating ($\theta$). 
Inserting this into Eq. \ref{inviscid}, we find the maximum growth
rate satisfies
\begin{align}\label{inviscid_max}
%  S_*^3 + \left[2(2-q)  - \frac{1}{\gamma \qtwo^2}\right]S_* - \frac{1}{\gamma
%    \qtwo^2 \beta} = 0.
  S_*^2 = \frac{1+\beta S_*}{Q^2\left(\theta + \beta\gamma S_*\right)}
  - 2(2-q).
\end{align}
Eq. \ref{kstar}---\ref{inviscid_max} imply $\p_\beta|K_*|,\,\p_\beta
S_*<0 $ for $\gamma>\theta$. Then as cooling becomes more rapid, the maximum 
growth rate increases, and the most unstable wavelength decreases. 
%show dK*/dbeta < 0 
%by diff K* wrt beta and diff S* wrt beta from the equations
%and elim dS*/d beta 
For $|S_*|\ll1$, Eq. \ref{inviscid_max} gives the simple solution 
\begin{align}
  \beta S_*\simeq \frac{1-2(2-q)Q^2\theta}{2(2-q)Q^2\gamma-1}.  
\end{align} 
Assuming the denominator is positive definite (see below for the case
when it vanishes),  then weakly-growing modes have
$S_*\propto\beta^{-1}$, so $\beta\gg1 $ but finite. That is, the disk
can be formally unstable for arbitrarily long cooling times.

%{\bf optional equation?}

@@

\subsubsection{$\theta = 0$}\label{theta0}
For $\beta$-cooling with $\theta=0$, the disk is unconditionally
unstable for finite $Q$, although instability occur on smaller scales
as $Q$ increases.  The $\beta\to0$ limit corresponds to a 
pressureless disk (not merely isothermal).

%for definiteness 
Let us consider a disk with 
\begin{align}\label{qcrit_def}
  \qtwo = \frac{1}{\sqrt{2\gamma(2-q)}}\equiv Q_{\mathrm{crit}},
\end{align} 
%{\bf KMK $\gamma = \theta$ from eq. 47? I'm missing something}
which is the condition for marginal stability in an adiabatic disk. 
How does finite cooling destabilize the disk?  
Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} with 
$\theta=0$, we find 
\begin{align}\label{sstar}
  S_*^3 = \frac{1}{\gamma Q_\mathrm{crit}^2 \beta}. 
\end{align}
The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
increases with decreasing $\beta$. 
%Instability formally occurs for
%arbitrarily long but finite cooling times. 
%Instability does not require the
%cooling time to be below some critical value.

We can define a characteristic cooling
time $\beta_*$ as that which removes pressure support against
self-gravity over the natural lengthscale in the problem, the
scale-height $H$. We thus set $|K_*|=1$ and find, for the Keplerian
disk,

\begin{align}\label{betastar}
  \beta_* = \frac{1}{\left(\sqrt{\gamma} - 1\right)^{3/2}}. 
\end{align}
This equation gives remarkably similar values 
of the cooling times below which 
numerical simulations show dynamical disk fragmentation 
\citep{gammie01,rice05,rice11}. These simulations employ the same beta  
cooling prescription with $\theta=0$, and determine the fragmentation
boundary, $\beta_c$, as a function of the adiabatic index $\gamma$.  
Table \ref{bstar_compare} shows rough agreement between $\beta_*$ and 
$\beta_c$.

@@

\subsubsection{$\theta = 1$}\label{theta1}
Standard beta cooling with $\theta=1$ corresponds to 
`thermal relaxation': the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case no
additional heat source need be invoked to define a steady state. From
Eq. \ref{stable_condition}, the instability condition is $Q<1$. This
is the same as the classic Toomre condition for an isothermal disk,  
which is formally obtained in the limit  $\beta\to 0$
(i.e. the disk temperature is constant in time). In this respect, a
fully irradiated disk, in that the equilibrium temperature is set
externally, behaves isothermally regardless of the cooling time.

%The most unstable wavenumber and growth
%rates vary weakly with $\beta$.  
% {\bf KMK: summarize boring result? not sure 
%what this simplifies to in comparison to other limits?}.
